---
title: "PeNSE 2024 · Bullying and Psychological Distress: Moderation Analysis"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
date: "2026-06-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r packages}
library(readxl)
library(lavaan)
library(semTools)
library(tidyverse)
library(psych)
library(knitr)
library(kableExtra)
```

```{r data}
df <- read_xlsx("pense_2024_clean.xlsx", na = c("null", ""))
describe(df)
colSums(is.na(df))
colMeans(is.na(df)) * 100
```

---

## 1. SEM Base Model

```{r sem-model}
sem_model <- "
  bullying               =~ bully_verbal + bully_relational + bully_physical
  psychological_distress =~ mh_worry + mh_sadness + mh_uncared + mh_irritable + mh_worthless
  psychological_distress ~ bullying
"
fit_sem <- sem(sem_model, data = df, ordered = TRUE, estimator = "WLSMV")
summary(fit_sem, fit.measures = TRUE, standardized = TRUE, rsquare = TRUE)
```

### Interpretation

The base SEM showed adequate fit (CFI = 0.952, RMSEA = 0.080, SRMR = 0.041). Bullying was a significant predictor of psychological distress (β = 0.525, p < .001), explaining 27.6% of its variance.

---

## 2. Moderator 1: School Support

```{r mod1}
fit_conf   <- sem(sem_model, data = df, ordered = TRUE, estimator = "WLSMV", group = "support")
fit_metric <- update(fit_conf, group.equal = "loadings")
fit_scalar <- update(fit_conf, group.equal = c("loadings", "thresholds"))
fit_var    <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances"))
fit_covar  <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances", "lv.covariances"))
teste      <- compareFit(fit_conf, fit_metric, fit_scalar, fit_var, fit_covar)
summary(teste)

fit_regression   <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances", "lv.covariances", "regressions"))
teste_moderation <- compareFit(fit_covar, fit_regression)
summary(teste_moderation)

subset(standardizedsolution(fit_covar), op == "~")

table(df$support)
lavInspect(fit_conf, "group.label")
```

### Measurement Invariance

Metric invariance was supported (ΔCFI = 0, ΔRMSEA = −0.006), as was scalar invariance (ΔCFI = 0, ΔRMSEA = +0.008), allowing meaningful group comparisons. Latent variable variances differed across groups (ΔCFI = 0, ΔRMSEA = −0.002).

### Moderation

School support significantly moderated the bullying–distress relationship (Δχ²(1) = 22.9, p < .001). Group order was confirmed via `lavInspect(fit_conf, "group.label")` = `"1" "0"`:

| Group | β |
|---|---|
| With support (1) | 0.520 |
| Without support (0) | 0.551 |

Although statistically significant given the large sample, the effect size difference (Δβ = 0.031) suggests that school psychological support had negligible protective value in this sample.

---

## 3. Moderator 2: Parental Understanding

```{r mod2}
fit_conf   <- sem(sem_model, data = df, ordered = TRUE, estimator = "WLSMV", group = "parent_understanding")
fit_metric <- update(fit_conf, group.equal = "loadings")
fit_scalar <- update(fit_conf, group.equal = c("loadings", "thresholds"))
fit_var    <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances"))
fit_covar  <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances", "lv.covariances"))
teste      <- compareFit(fit_conf, fit_metric, fit_scalar, fit_var, fit_covar)
summary(teste)

fit_regression   <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances", "lv.covariances", "regressions"))
teste_moderation <- compareFit(fit_covar, fit_regression)
summary(teste_moderation)

subset(standardizedsolution(fit_covar), op == "~")

table(df$parent_understanding)
lavInspect(fit_conf, "group.label")
```

### Measurement Invariance

Metric invariance was supported (ΔCFI = −0.002, ΔRMSEA = −0.006), as was scalar invariance (ΔCFI = 0, ΔRMSEA = +0.010), allowing meaningful group comparisons. Latent variable variances were not invariant across groups (ΔCFI = −0.014), indicating differences in construct dispersion.

### Moderation

Parental understanding significantly moderated the bullying–distress relationship (Δχ²(2) = 3077.9, p < .001). Group order was confirmed via `lavInspect(fit_conf, "group.label")` = `"2" "1" "0"`:

| Group | β |
|---|---|
| Most times/Always (2) | 0.521 |
| Rarely/Sometimes (1) | 0.443 |
| Never (0) | **0.804** |

Adolescents whose parents never understand their problems showed a substantially stronger bullying–distress effect (β = 0.804), nearly double that of the partial understanding group (β = 0.443). This suggests parental understanding acts as a meaningful protective factor, even when only partial.

---

## 4. Moderator 3: Peer Kindness

```{r mod3}
fit_conf   <- sem(sem_model, data = df, ordered = TRUE, estimator = "WLSMV", group = "peer_kindness")
fit_metric <- update(fit_conf, group.equal = "loadings")
fit_scalar <- update(fit_conf, group.equal = c("loadings", "thresholds"))
fit_var    <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances"))
fit_covar  <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances", "lv.covariances"))
teste      <- compareFit(fit_conf, fit_metric, fit_scalar, fit_var, fit_covar)
summary(teste)

fit_regression   <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances", "lv.covariances", "regressions"))
teste_moderation <- compareFit(fit_covar, fit_regression)
summary(teste_moderation)

subset(standardizedsolution(fit_covar), op == "~")

table(df$peer_kindness)
lavInspect(fit_conf, "group.label")
```

### Measurement Invariance

Metric invariance was supported (ΔCFI = −0.003, ΔRMSEA = −0.004), as was scalar invariance (ΔCFI = +0.002, ΔRMSEA = +0.008), allowing meaningful group comparisons. Latent variable variances were not invariant across groups (ΔCFI = −0.007).

### Moderation

Peer kindness significantly moderated the bullying–distress relationship (Δχ²(2) = 2551.9, p < .001). Group order was confirmed via `lavInspect(fit_conf, "group.label")` = `"2" "1" "0"`:

| Group | β |
|---|---|
| Most times/Always (2) | 0.478 |
| Rarely/Sometimes (1) | 0.506 |
| Never (0) | **0.870** |

A clear dose-response pattern emerged: the less kindness experienced from peers, the stronger the impact of bullying on psychological distress. Adolescents who never experienced peer kindness showed the highest effect (β = 0.870), nearly twice that of those with consistent peer support (β = 0.478).

---

## 5. Moderator 4: Number of Close Friends

```{r mod4}
fit_conf   <- sem(sem_model, data = df, ordered = TRUE, estimator = "WLSMV", group = "n_friends")
fit_metric <- update(fit_conf, group.equal = "loadings")
fit_scalar <- update(fit_conf, group.equal = c("loadings", "thresholds"))
fit_var    <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances"))
fit_covar  <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances", "lv.covariances"))
teste      <- compareFit(fit_conf, fit_metric, fit_scalar, fit_var, fit_covar)
summary(teste)

fit_regression   <- update(fit_conf, group.equal = c("loadings", "thresholds", "lv.variances", "lv.covariances", "regressions"))
teste_moderation <- compareFit(fit_covar, fit_regression)
summary(teste_moderation)

subset(standardizedsolution(fit_covar), op == "~")

table(df$n_friends)
lavInspect(fit_conf, "group.label")
```

### Measurement Invariance

Metric invariance was supported (ΔCFI = −0.001, ΔRMSEA = −0.005). Scalar invariance showed a borderline pattern (ΔCFI = +0.001, ΔRMSEA = +0.007). Latent variable variances differed across groups (ΔCFI = −0.004).

### Moderation

Number of close friends significantly moderated the bullying–distress relationship (Δχ²(1) = 689.5, p < .001). Group order was confirmed via `lavInspect(fit_conf, "group.label")` = `"1" "0"`:

| Group | β |
|---|---|
| Has friends (1+) | 0.520 |
| No friends | **0.822** |

Adolescents with no close friends showed a substantially stronger bullying–distress effect (β = 0.822 vs β = 0.520). Having even one close friend appears to confer meaningful protection against the psychological impact of bullying.

---

## 6. Summary

```{r summary-table}
summary_df <- data.frame(
  Moderator       = c("School Support", "Parental Understanding", "Peer Kindness", "N Friends"),
  `Delta Chi2`    = c(22.9, 3077.9, 2551.9, 689.5),
  df              = c(1, 2, 2, 1),
  p               = c("<.001", "<.001", "<.001", "<.001"),
  `Beta protected`   = c(0.520, 0.443, 0.478, 0.520),
  `Beta unprotected` = c(0.551, 0.804, 0.870, 0.822),
  `Delta Beta`    = c(0.031, 0.361, 0.392, 0.302),
  Protective      = c("❌ Negligible", "✅ Strong", "✅ Strong", "✅ Strong")
)

kable(summary_df, digits = 3, caption = "Moderation Effects Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

Three of four moderators demonstrated meaningful protective effects against the bullying–psychological distress relationship. Peer kindness showed the largest differential (Δβ = 0.392), followed by parental understanding (Δβ = 0.361) and number of close friends (Δβ = 0.302). School psychological support, while statistically significant, showed negligible practical effect (Δβ = 0.031), possibly reflecting limitations in the reach or quality of school-based services captured in this dataset.
